---
title: 'DS 202 - Lab 4: Tidying and Joining'
subtitle: "Thursday 10/24/2019 by 10 pm"
author: "Jennie Huxford"

date: "10/17/2019"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Format for submission

- We will check whether or not your lab results are reproducible - i.e. we will knit your `team.Rmd` file. 
- You will also be graded on **style**. Style means is your lab *clear, easy to read, and flows logically*? This lab does **not** need to be in report format but it should be neatly organized with appropriate `R` chunks. 
-  <span style="color:red">Each plot at **minimum** must have labels for both axes and a title. If you don't have these components, you will not get full credit for your answers. Each plot ideally should also come with a description. </span> 

## Submission
Your git repo will contain the following files: `team.Rmd` and `team.html` containing the collaborated work.

- Both teammates need to modify `team.Rmd`. `git log team.Rmd` should show commit information from both teammates.
- To submit: 

  1. Push `team.Rmd` and `team.html` to the owner's git repo.

  2. **Submit both `team.Rmd` and `team.html` to Canvas as well**. In your submission, include: 
    - The owner's GitHub username 
    - The partner's GitHub username 
    - Repo name (should be `ds202_lab4`)
- Due date: Thursday Oct. 24 by 10 pm. 



## Background
An excerpt of the individual player data of the Cyclone football team is available from Canvas. Dowload the [excel file](cyclonesFootball2018.xlsx) and put the file next to your Rmd files. 

Some abbreviations for the column names can be found in the Glossaries sheet in the Excel file.

**References:**

- [Iowa State University Athelete](https://cyclones.com/index.aspx)
- [A Beginner's Guide to American Football](https://www.youtube.com/watch?v=3t6hM5tRlfA) for the rules of American football.

## Loading Data: Cyclone Football 

The Cyclone football data involves multiple Excel sheets and is in a somewhat messy form. The following code loads the Defensive sheet into R. **For this lab, also import the Offensive and the Biography sheets.**

```{r, include= FALSE}
library(reshape2)
library(dplyr)
library(tidyr)
```

```{r}
library(readxl)
defense <- read_excel('cyclonesFootball2018.xlsx', sheet='Defensive')
str(defense)
```

```{r,echo=FALSE,results='hide'}
offense <- read_excel('cyclonesFootball2018.xlsx', sheet='Offensive')
bio <- read_excel('cyclonesFootball2018.xlsx', sheet='Biography')
```

## Part one: Cleaning data
The three Excel sheets are loaded into R but they are not useful to us before some cleaning. Manipulate the three imported data frames, so that the cleaned data satisfy the following:

1. The player names and opponent names are factors.
```{r}
offense$Name <- as.factor(offense$Name)
offense$Opponent_Opponent <- as.factor(offense$Opponent_Opponent)
defense$Name <- as.factor(defense$Name)
defense$Opponent_Opponent <- as.factor(defense$Opponent_Opponent)
bio$Name <- as.factor(bio$Name)
```

2. All offensive and defensive statistics are numerical. The `Weight` column in Biography is also numeric. (Hint: look at `?mutate_at` if you want to save writing some code). 

```{r}
offense <- offense %>% mutate_at(c('Receiving_REC', 'Receiving_TD', 'Receiving_YDS', 'Rushing_ATT', 'Rushing_YDS', 'Rushing_TD', 'Passing_CMP-ATT', 'Passing_YDS', 'Passing_TD', 'Passing_INT'), as.numeric)
```
```{r}
defense <- defense %>% mutate_at(c('Tackles_Solo', 'Tackles_ASST', 'Tackles_TFL', 'Tackles_Sack', 'Turnover_FF', 'Turnover_FR', 'Turnover_INT', 'Pass_QBH', 'Pass_PB'), as.numeric)
```
```{r}
bio <- bio %>% mutate_at(c('Weight'), as.numeric)
```

3. Separate the `Height` column into two columns. Make sure the newly created columns are numeric. Then create one column that represents `Height` (Hint: Use a unit that works well). Drop any irrelavant columns from this process.
```{r}
bio <- bio %>% separate(Height, into = c("Height (feet)", "Height (inches)")) %>% mutate_at(c('Height (feet)', 'Height (inches)'), as.numeric)
head(bio)
```
```{r}
bio<- bio %>% unite("Height", c("Height (feet)", "Height (inches)"), sep = "'", remove = TRUE)
```
4. By cleaning the three data frames in R, at the end of this step you should have the following three clean data frames: `defClean`, `offClean`, and `bioClean`. Print the structures of these data frames. **Make sure your data types make sense for each variable.**
```{r}
defClean <- data.frame(defense)
head(defClean)
offClean <- data.frame(offense)
head(offClean)
bioClean <- data.frame(bio)
head(bioClean)
```

## Part two: Tidying (A)
For the following questions, work on the cleaned data frames `defClean`, `offClean`, and `bioClean`.

1. Reformat the `defClean` data frame into a tidy format using `gather`. The type of statistic (Tackles_Solo, Tackles_ASST, etc) is added as a **new key column named `Statistic`.** 

```{r}
defClean<- defClean %>% gather(key = Statistic , value = value, 3:11)
head(defClean)
```

2. Create a histogram for each defensive statistic (make sure to explore different binwidths). Use `facet_wrap` to draw all histograms in a single plot. What defensive skills are rare?
```{r}
library(ggplot2)
```
```{r}
ggplot(defClean, aes(value)) + geom_histogram(binwidth = 0.5) + facet_wrap(~Statistic)
```


The rare defensive skills are turnovers from fumble returns, forced fumbles and interceptions, along with sacks. These defensive skills appear to have a lot fewer occurances than the 5 skills based on the graphs above. 

3. Did ISU have better defense against West Virginia or Kansas State? Answer this question by creating a scatterplot with x-axis being the number of solo tackles (by each player) against West Virginia, and y-axis being that against Kansas State. A large number of solo tackles indicates good defense. 

      (**Hint**: From step 1 we have gathered our `defClean` into 4 columns. \
       Now we only want to focus on solo tackles (hint...`filter`). 
      Look at your data structure. What form do we need our data to have in order to easily answer this question? (hint...`spread`)). 

```{r}
solotackles <- defClean %>% filter(Statistic == "Tackles_Solo", Opponent_Opponent %in% c("Kansas State", "West Virginia"))
solotackles <- solotackles %>% spread(key = Opponent_Opponent, value = value)
na.omit(solotackles)
```

```{r}
solotackles$ks <- solotackles$`Kansas State`
solotackles$wv <- solotackles$`West Virginia`
ggplot(solotackles, aes(ks, wv)) + geom_point() + ggtitle("Kansas State Solo Tackles vs. West Virginia Solo Tackles") +
  xlab("Kansas State Solo Tackles") +
  ylab("West Virginia Solo Tackles")
```


Kansas State appears to have had a better defense than West Virginia did against ISU. Kansas State had 7 players with over 2.5 solo tackles whereas West Virginia only had two players with over 2.5 solo tackles. 

## Part two: Tidying  (B)
Now let's turn our attention to the player biography `bioClean`.

1. Separate the `Hometown` column into two columns, namely the city and the state. Print the first few rows of these two columns. (Hint: look at the `sep=` argument of `separate`)

```{r}
bioClean <- bioClean %>% separate(Hometown, into = c("City", "State"), sep = ",")
citystate<- bioClean %>% select(City, State)
head(citystate)
```

2. How many players are there from each state?

```{r}
table(bioClean$State)
```


## Part three: Joining data frames

1. Find out where the offensive players are from by joining the offensive data and player biography data appropriately.
```{r}
offbio <- left_join(offClean, bioClean, by = "Name")
offbio_distinct <- offbio %>% distinct(Name, State) %>% select(Name, State)
table(offbio_distinct$State)
```


2. Is there a relationship between the weight of a player and the receiving yards (`Receiving_YDS`)?

```{r}
ggplot(offbio, aes(Weight, Receiving_YDS)) + geom_point() + ggtitle("Weight and Receiving Yards") +
  xlab("Weight") +
  ylab("Receiving Yards")
```


Looking at the graph above there does not appear to be an obvious strong relationship between weight of the player and receiving yards. However it does look as if the lower weights tend to have higher recieving yards, except for the player at 225 lbs.

3. Explore more. Raise a question that can be answered only by looking at two or more data frames. Then answer that question.

Is the weight of a player associated with the number of sacks they have? 
```{r}
defbio<- left_join(defClean, bioClean, by = "Name")
sacks <- defbio %>% filter(Statistic == "Tackles_Sack")
```

```{r}
ggplot(sacks, aes(Weight, value)) + geom_point() + ggtitle("Weight and # of Sacks") +
  xlab("Weight") +
  ylab("# of Sacks")
```


Based on the graph above the weight of a player does not seem to be associated with the number of sacks a player has. Most of the points are around zero which makes a lot of sense because a player getting a sack is a pretty rare occurance. We also see from the graph that there are only two players with two sacks, but about 14 with one sack.
## Tips
- Agree on a method to clean the dataset with your teammate. Both teammates can then work on the same cleaned version for downstream analysis.
- Start early to allow more time to collaborate.
- Plan ahead to prevent merge conflict: Teammates should modify the `team.Rmd` file in turn, rather than concurrently. 

